---
title: "An R Refresher"
subtitle: "Essential Programming Skills for Longitudinal Data Analysis"
author: "[Dr Clemens Jarnach](https://clemensjarnach.github.io)"
affiliation: "Oxford Martin School, University of Oxford"
date: "03-18-2026"
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Generate panel_data.csv if it does not already exist
if (!file.exists("files/panel_data.csv")) {
  source("simple_panel_data.r")
}

panel_data <- readr::read_csv("files/panel_data.csv", show_col_types = FALSE)
```

Before we run models, we need to ensure you're comfortable with R itself. This chapter introduces the R programming concepts you will need throughout the workshop. If you are already comfortable with R, treat it as a quick reference. If you are new to R, work through it carefully — the material here underpins everything that follows.

::: callout-tip
## Running code

All code blocks in this workshop can be run interactively. Open the `.qmd` source file in RStudio, place your cursor inside a code chunk, and press **Ctrl+Enter** (Windows/Linux) or **Cmd+Enter** (Mac) to run a single line, or **Ctrl+Shift+Enter** to run the entire chunk.
:::

------------------------------------------------------------------------

## Basic Operations and Assignment

R can be used as a calculator, but its real power lies in storing and manipulating values. The assignment operator `<-` binds a value to a name in the current environment.

```{r}
1 + 3   
# arithmetic evaluation
```

```{r}
2^8             # exponentiation
```

```{r}
17 %% 5         # modulo (remainder)
```

```{r}
17 %/% 5        # integer division
```

```{r}

a <- 9          # assignment: bind 9 to the name 'a'
```

```{r}
sqrt(a)         # apply a function to an object
```

```{r}
b <- sqrt(a)
```

```{r}
a == b          # logical comparison: are they equal?
```

```{r}
a != b          # not equal
```

```{r}
a > 5           # greater than
```

```{r}
ls()            # list all objects in the current environment
```

::: callout-note
R is case-sensitive: `A` and `a` are different objects. Object names can contain letters, digits, `.` and `_`, but must start with a letter or `.`.
:::

------------------------------------------------------------------------

## Data Types

Every value in R has a type. The most common scalar types are:

```{r}
class(3.14)         # numeric (double)
class(42L)          # integer (note: I used the L suffix to tell R this is an integer)
class("hello")      # character
class(TRUE)         # logical
class(2 + 3i)       # complex

# Type coercion
as.numeric("3.14")
as.integer(42)
as.character(100)
as.logical(0)       # 0 is FALSE; anything non-zero is TRUE
is.na(NA)           # test for missing value
```

------------------------------------------------------------------------

## Vectors

A vector is the fundamental data structure in R — an ordered collection of values of the **same type**. Almost all operations in R are vectorised, meaning they apply element-wise without explicit loops.

```{r}
x <- c(1, 3, 5)          # combine values into a vector
y <- c("one", "three", "five")
```

```{r}
# Indexing (1-based)
x[1]                      # first element
x[c(1, 3)]                # first and third elements
x[-2]                     # all except the second
```

```{r}
# Sequences
a <- 1:10
b <- seq(from = 0, to = 1, by = 0.25)
c_rep <- rep(c(1, 2), times = 3)
c_rep
```

```{r}
# Vectorised operations — applied element-wise
x * 2
x + c(10, 20, 30)
```

```{r}
# Logical operations on vectors
x > 2
any(x > 2)                # is any element > 2?
all(x > 2)                # are all elements > 2?
which(x > 2)              # indices where condition is TRUE
```

```{r}
# Common summary functions
length(x)
sum(x)
mean(x)
sd(x)
min(x); max(x)
```

------------------------------------------------------------------------

## Matrices

A matrix is a two-dimensional vector: all elements share the same type, and values are arranged in rows and columns. In longitudinal data analysis, matrices appear in covariance and correlation structures, variance-component decompositions, and balanced panel layouts — understanding matrix indexing is useful for inspecting model objects and constructing derived quantities.

```{r}
m <- matrix(1:25, nrow = 5, ncol = 5)
m
```

```{r}
# Element access
m[1, 2]                   # row 1, column 2
m[1, ]                    # entire first row
m[, 2]                    # entire second column
m[2:3, 3:5]               # submatrix
```

```{r}
# Dimensions
nrow(m); ncol(m)
dim(m)
```

```{r}
# Matrix operations — useful for covariance and correlation matrices in panel diagnostics
t(m)                      # transpose
m %*% t(m)                # matrix multiplication (not element-wise!)
diag(m)                   # diagonal elements
```

```{r}
# Compute a correlation matrix — useful for checking multicollinearity among panel predictors
vars <- data.frame(
  gdp_growth = c(1.2, 2.3, 1.8, 3.1, 0.9, 2.7),
  trade_open  = c(0.4, 0.9, 0.6, 1.1, 0.3, 0.8),
  population  = c(10,  12,  11,  14,  9,   13)
)
cor(vars)                 # pairwise correlations between panel-level variables
```

------------------------------------------------------------------------

## Lists

Lists are R's most flexible container: each element can hold any object of any type or length. Many R functions return lists (model output from `plm`, `lme4`, or `fixest`, etc.), so you need to know how to navigate them.

```{r}
person <- list(
  name    = "Alice",
  age     = 34,
  scores  = c(88, 92, 79),
  active  = TRUE
)
```

```{r}
# Access by name
person$name
person[["age"]]
```

```{r}
# Access by position
person[[3]]               # third element (the scores vector)
person[[3]][2]            # second score
```

```{r}
# Inspect structure
str(person)
length(person)
names(person)
```

------------------------------------------------------------------------

## Data Frames

A data frame is the standard rectangular data structure in R: a list of vectors of equal length, where each vector is a column. Think of it as a spreadsheet or database table. Longitudinal datasets are naturally represented as data frames, where each row is an observation for a given unit (individual, country, firm) at a given time point.

```{r}
df <- data.frame(
  id     = 1:5,
  name   = c("Alice", "Bob", "Clara", "David", "Eva"),
  degree = c(3, 7, 2, 5, 4),
  active = c(TRUE, TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

df
```

```{r}
# Column access
df$name
df[["degree"]]
```

```{r}
# Row and column indexing
df[1, ]                   # first row
df[, 3]                   # third column
df[df$active, ]           # rows where active == TRUE
df[df$degree > 3, c("name", "degree")]
```

```{r}
# Useful inspection functions
nrow(df); ncol(df)
head(df, 3)
str(df)
summary(df)
```

------------------------------------------------------------------------

## Control Flow

### Conditionals

```{r}
x <- 15

if (x > 10) {
  cat("x is greater than 10\n")
} else if (x == 10) {
  cat("x is exactly 10\n")
} else {
  cat("x is less than 10\n")
}

# ifelse() — vectorised conditional
scores <- c(55, 72, 48, 91, 60)
ifelse(scores >= 60, "pass", "fail")
```

### Loops

In R, explicit loops are often avoidable thanks to vectorisation, but they are useful for iterative tasks and are worth knowing.

```{r}
# for loop
for (i in 1:5) {
  cat("Iteration:", i, "\n")
}
```

```{r}
# while loop
count <- 0
while (count < 3) {
  count <- count + 1
  cat("count is", count, "\n")
}
```

```{r}
# break and next
for (i in 1:10) {
  if (i == 4) next       # skip this iteration
  if (i == 7) break      # exit the loop
  cat(i, "")
}
```

### Apply Functions

The `apply` family offers a concise way to apply a function over a vector, list, or matrix — often replacing explicit loops.

```{r}
m <- matrix(1:12, nrow = 3)
```

```{r}
apply(m, 1, sum)          # row sums
apply(m, 2, mean)         # column means
```

```{r}
nums <- list(a = 1:4, b = 5:10, c = 11:15)
sapply(nums, mean)        # returns a named vector
lapply(nums, length)      # returns a list
```

------------------------------------------------------------------------

## Writing Functions

Functions are the building blocks of reusable code. When you find yourself repeating the same operations, write a function.

```{r}
# Basic function definition
greet <- function(name, greeting = "Hello") {
  paste(greeting, name)
}

greet("Alice")
greet("Bob", greeting = "Welcome")
```

```{r}
# A more practical example: normalise a vector to [0, 1]
normalise <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

```{r}
gdp_growth <- c(1.3, 3.7, 0.2, 4.9, 2.4)
normalise(gdp_growth)
```

```{r}
# Functions can return multiple values via a list
describe <- function(x) {
  list(n = length(x), mean = mean(x), sd = sd(x), range = range(x))
}

describe(gdp_growth)
```

------------------------------------------------------------------------

## Data Wrangling with the Tidyverse

The [tidyverse](https://www.tidyverse.org) is a collection of R packages designed around a consistent philosophy of tidy data and readable code. The core package for data manipulation is **dplyr**, which provides a small set of verbs that cover the vast majority of data-wrangling tasks.

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

### The Pipe Operator

The pipe `|>` (base R, since 4.1) or `%>%` (magrittr/tidyverse) passes the result of one expression as the first argument of the next. It allows you to write a sequence of transformations in the order they happen, which is much easier to read than nested function calls.

```{r}
# Without pipe — read inside out
round(mean(c(1, 2, 3, 4, 5)), digits = 2)
```

```{r}
# With pipe — read left to right
c(1, 2, 3, 4, 5) |> mean() |> round(digits = 2)
```

### A Working Dataset

We'll use a small cross-sectional dataset of survey respondents, representative of the kind of individual-level data you might encounter in longitudinal research.

```{r}
persons <- tibble(
  id         = 1:8,
  name       = c("Alice","Bob","Clara","David","Eva","Frank","Grace","Hugo"),
  country    = c("UK","France","UK","Germany","France","UK","Germany","France"),
  age        = c(32, 45, 28, 51, 38, 23, 47, 35),
  income     = c(42000, 68000, 35000, 72000, 55000, 28000, 61000, 49000)
)
```

```{r}
persons
```

### filter() — Keep Rows

```{r}
# Keep only UK respondents
persons |> filter(country == "UK")
```

```{r}
# Multiple conditions
persons |> filter(country == "France", age > 40)
```

```{r}
# Using %in% for multiple values
persons |> filter(country %in% c("UK", "Germany"))
```

### select() — Choose Columns

```{r}
persons |> select(name, country, income)
```

```{r}
# Drop a column with -
persons |> select(-id)
```

```{r}
# Rename while selecting
persons |> select(name, cntry = country, annual_income = income)
```

### mutate() — Create or Modify Columns

```{r}
persons |>
  mutate(
    older       = age >= 40,
    income_z    = (income - mean(income)) / sd(income),   # standardise
    label       = paste0(name, " (", country, ")")
  )
```

### arrange() — Sort Rows

```{r}
persons |> arrange(desc(income))       # highest income first
persons |> arrange(country, age)       # sort by two columns
```

### summarise() and group_by() — Aggregation

```{r}
# Overall summary
persons |>
  summarise(
    n           = n(),
    mean_income = mean(income),
    sd_income   = sd(income),
    max_age     = max(age)
  )
```

```{r}
# Summary by group — useful for comparing units in panel data
persons |>
  group_by(country) |>
  summarise(
    n           = n(),
    mean_income = mean(income),
    mean_age    = mean(age)
  ) |>
  arrange(desc(mean_income))
```

### Joining Tables

Joins combine two data frames on a shared key column — essential in longitudinal work when merging panel waves with time-invariant attributes, or linking datasets from different sources.

```{r}
# Panel waves: repeated observations per person across two years
waves <- tibble(
  id   = c(1, 1, 2, 2, 3, 3, 4, 4),
  year = c(2020, 2021, 2020, 2021, 2020, 2021, 2020, 2021),
  income_obs = c(40000, 42000, 65000, 68000, 33000, 35000, 70000, 72000)
)
```

```{r}
# Attach time-invariant person attributes to the panel
waves |>
  left_join(persons |> select(id, name, country),
            by = "id") |>
  arrange(id, year)
```

### Reshaping Data

```{r}
# Wide to long (pivot_longer) — core operation when converting panel data to long format
persons |>
  select(name, age, income) |>
  pivot_longer(cols = c(age, income),
               names_to  = "variable",
               values_to = "value")
```

```{r}
# Long to wide (pivot_wider) — useful when reshaping repeated measures to one row per unit
waves |>
  select(id, year, income_obs) |>
  pivot_wider(names_from = year, values_from = income_obs,
              names_prefix = "income_")
```

------------------------------------------------------------------------

## Reading and Writing Data

```{r}
#| eval: false

# CSV files
df <- read_csv("data/panel_data.csv")       # tidyverse (recommended)
df <- read.csv("data/panel_data.csv")       # base R
```

```{r}
#| eval: false
write_csv(df, "data/panel_data_clean.csv")
```

```{r}
#| eval: false
# Excel files
library(readxl)
df <- read_excel("data/data.xlsx", sheet = 1)
```

```{r}
#| eval: false
# SPSS, Stata, SAS
library(haven)
df <- read_spss("data/survey.sav")
df <- read_dta("data/survey.dta")
```

```{r}
#| eval: false
# R's native format
saveRDS(df, "data/df.rds")
df <- readRDS("data/df.rds")
```

------------------------------------------------------------------------

## Quick Reference

| Task                   | Function                                |
|------------------------|-----------------------------------------|
| Assign a value         | `x <- value`                            |
| Create a vector        | `c(1, 2, 3)`                            |
| Create a sequence      | `1:10`, `seq(0, 1, 0.1)`                |
| Create a matrix        | `matrix(data, nrow, ncol)`              |
| Create a data frame    | `data.frame()` / `tibble()`             |
| Check an object's type | `class()`, `typeof()`                   |
| Check dimensions       | `dim()`, `nrow()`, `ncol()`, `length()` |
| Inspect structure      | `str()`, `summary()`, `head()`          |
| Filter rows            | `filter()`                              |
| Select columns         | `select()`                              |
| Create columns         | `mutate()`                              |
| Sort rows              | `arrange()`                             |
| Aggregate              | `group_by()` + `summarise()`            |
| Join tables            | `left_join()`, `inner_join()`           |
| Reshape wide → long    | `pivot_longer()`                        |
| Reshape long → wide    | `pivot_wider()`                         |
